---
title: "COVID Vaccination Project"
author: "Adrian Cornejo and Victor Guillera"
date: "11/4/2021"
output: pdf_document
---

# Milestone 2

Section 1 

This data set will relate to our question if being a homeowner status and 
available housing impacts vaccination rates per CA county. While not a perfect 
representation of SES of cases, this data set will help us elucidate if there is
a trend via linear regression between the homeowner-related to renter rates and
the vaccination rate per county.

We have two data sets, one is of COVID-19 vaccination progress throughout 
CA provided by the CDPH following vaccine dosage by zip code from 1/5/21 
to about present 9/14/21, and the other is demographic info in CA across 
all counties from the US census data for 2014 to 2019.
Below, we read in our libraries, data sets, explore our data sets, 
and then proceed to clean the data sets.



```{r}
library(readr)
library(knitr)
library(tibble)
library(dplyr)
library(stringr)
countydem <- read.csv("ca_county_demographics.csv")
covidvax <- read.csv("cov_vax_admin.csv")

```


To import the csv files, we used read_csv from base R and will 
inspect our imported files first before we clean our data set.
Libraries we loaded in were readr, dplyr, and tibble. We are going to pull 
variables "persons_fully_vaccinated", "persons_partially_vaccinated", 
"zip_code_tabulation_area", "age12_plus_population", and "county" from the
covidvax csv file. We will pull "name", "renter_occ","owner_occ", "hse_units" 
and "vacant" variables.
```{r, results = "hide"}
str(covidvax)
head(covidvax)
nrow(covidvax)
ncol(covidvax)
typeof(covidvax)
class(covidvax)
```

```{r, results = "hide"}
distinct(covidvax, as_of_date, .keep_all = F)
```

The structure of the "cov_vax_admin" file is a data frame with 11 columns and 
65,268 rows. The data frame contains integer, number, and character data types 
and contains a redacted column that is inaccessible. There are also a total of 
27 unique dates, each 7 weeks apart (weekly results). 


```{r, results = "hide"}
str(countydem)
head(countydem)
nrow(countydem)
ncol(countydem)
typeof(countydem)
class(countydem)
```
The structure of "countydem" reveals a data frame containing 
23 columns and 58 rows with integer, character and number data types.

\newpage

```{r, results = "hide" } 
#Cleaning the data by removing unnecessary columns, rearranging the data,
#and removing "N/A" data points.

#COVID Vaccination Rates
covidvax2 <- select(covidvax, "county","zip_code_tabulation_area", "as_of_date",
                    "age12_plus_population", "persons_fully_vaccinated",
                    "persons_partially_vaccinated")

names(covidvax2) <- c("county", "zip_code", "date", "total_pop_over_12", 
                      "fully_vaccinated", "partially_vaccinated")


covidvax2[is.na(covidvax2)] <- 0

#Using the function str() we see that the selected varibales in covidvax2 
#are chr, int, chr, num, num 
#We may need to convert the dates to numeric using the as.numeric() function 
#to determine and compare vaccination rate. All other column types should be 
#fine as is. 
#Converting NA cells to 0 we allow ourselves to manipulate and describe entire 
#numeric columns.
```
```{r, eval=F}
#Descriptions
distinct(covidvax2, county, .keep_all = F)
distinct(covidvax2, date, .keep_all = F)
distinct(covidvax2, zip_code, .keep_all = F)

covidvax2 %>% group_by(zip_code) %>%
  summarise(fullyvax_range = range(fully_vaccinated), partiallyvax_range = range(partially_vaccinated))

covidvax2 %>% 
  summarise(total_pop_range = range(total_pop_over_12), total_pop_median = median(total_pop_over_12))
```

```{r, results = "hide"}
#County Data 

countydem2 <- select(countydem, "name", "renter_occ","owner_occ", 
                     "vacant", "hse_units")

names(countydem2) <-c("county_name","renters","owners",
                      "vacant_housing", "total_housing_units")

countydem2[is.na(countydem2)] <- 0
```

Using the function str() we see that the selected varibales in 
countydem2 are chr, int, int, int, int. All column types should be fine as is 
for our study question. Converting NA cells to 0 we allow ourselves to 
manipulate and describe entire numeric columns. We also changed column names to 
allow for easier/obvious verbiage.

\newpage

```{r, results = "hide"}

#Descriptions

distinct(covidvax2, county, .keep_all = F)
distinct(covidvax2, date, .keep_all = F)
distinct(covidvax2, zip_code, .keep_all = F)

```

```{r, results = "hide"}

covidvax2 %>% group_by(zip_code) %>% 
  summarise(fullyvax_range = range(fully_vaccinated), partiallyvax_range = range(partially_vaccinated))

``` 
Population Range and Median for COVID vaccinations for those over 12 years old.
```{r, results = "hide"}
covidvax2 %>% summarise(total_pop_range = range(total_pop_over_12),
                        total_pop_median = median(total_pop_over_12))

#Using the function str() we see that the selected variables 
#in countydem2 are chr, int, int, int, int 
#All column types should be fine as is for our study question
#Converting NA cells to 0 we allow ourselves to manipulate and describe 
#entire numeric columns.
#Changed column names to easier/obvious verbiage
```
#Descriptions
```{r, results = "hide"}
distinct(countydem2, county_name, .keep_all = F)

countydem2 %>% summarise(renters_mean = mean(renters),
                         owners_mean = mean(owners), 
                         vacant_mean = mean(vacant_housing), total_mean = mean(total_housing_units))

countydem2 %>% summarise(renters_range = range(renters), 
                         owners_range = range(owners), 
                         vacant_range = range(vacant_housing), total_range = range(total_housing_units))
```

#Below, is the minimum, first quartile, median, third quartile, and max values.

```{r, results = "hide"}
fivenum(covidvax2$total_pop_over_12)
fivenum(covidvax2$fully_vaccinated)
fivenum(covidvax2$partially_vaccinated)
```


```{r, results = "hide"}


#Descriptions

#We get the mean & range for all numeric and integer vectors in countydem2 below

distinct(countydem2, county_name, .keep_all = F)

``` 

```{r, results = "hide"}
countydem2 %>%
  summarise(renters_mean = mean(renters), 
            owners_mean = mean(owners), 
            vacant_mean = mean(vacant_housing), 
            total_mean = mean(total_housing_units))

countydem2 %>% 
  summarise(renters_range = range(renters), 
                         owners_range = range(owners), 
                         vacant_range = range(vacant_housing), 
                         total_range = range(total_housing_units))

```
Below, we get the five number summary for data of interest in countydem2.
```{r}
fivenum(countydem2$renters)
fivenum(countydem2$owners)
fivenum(countydem2$vacant_housing)
fivenum(countydem2$total_housing_units)

```
\newpage

# Milestone 3

```{r, results = "hide"}

#Covid Vaccination by County data manipulation (covidvax files)

unique(covidvax2$county)

covidvax2 <- covidvax2 %>% filter(county!="0")

unique(covidvax2$date)
```
```{r, results = "hide"}

covidvax3 <- covidvax2 %>% 
  mutate(month = case_when(
    str_detect(date, "-01-") == T ~ "January",
    str_detect(date, "-02-") == T ~ "February",
    str_detect(date, "-03-") == T ~ "March",
    str_detect(date, "-04-") == T ~ "April",
    str_detect(date, "-05-") == T ~ "May",
    str_detect(date, "-06-") == T ~ "June",
    str_detect(date, "-07-") == T ~ "July",
    str_detect(date, "-08-") == T ~ "August",
    str_detect(date, "-09-") == T ~ "September"
  ))

unique(covidvax3$month)

```
`
```{r, results = "hide"}

covidvax4 <- covidvax3 %>% 
  group_by(county,month)%>%
  mutate(county_total_pop_over12 = sum(total_pop_over_12)) %>%
  mutate(county_fully_vaccinated = sum(fully_vaccinated)) %>%
  mutate(county_partially_vaccinated = sum(partially_vaccinated))
  

covidvax4 <- covidvax4 %>%
  select(-"zip_code",-"total_pop_over_12",-"fully_vaccinated",
         -"partially_vaccinated")

#Ensuring we get the last date for each month's final total vaccinations

unique(covidvax4$date)

covidvax4 <- covidvax4 %>% filter(date == "2021-01-26"| date == "2021-02-23" |
                                    date == "2021-03-30" |
                                    date == "2021-04-27" |
                                    date == "2021-05-25" |
                                    date == "2021-06-29" |
                                    date == "2021-07-27" |
                                    date == "2021-08-31" |
                                    date == "2021-09-14")

```

```{r, results = "hide"}

#Removing duplicate columns after combining the zip codes for each county
#and making column totals for each . Then we check to ensure we have 58 counties
#for each month.

covidvax5 <- covidvax4 %>% distinct(county,month, .keep_all=T)

covidvax5 %>% group_by(month) %>% summarise(n())
```
```{r, results = "hide"}
covidvax5 <- covidvax5 %>%
mutate(county_percent_fullvax =
(county_fully_vaccinated)/(county_total_pop_over12)) %>%
mutate(county_percent_partialvax =
(county_partially_vaccinated)/(county_total_pop_over12))

#Checking that our percentages make sense.
summary(covidvax5$county_percent_fullvax)
summary(covidvax5$county_percent_partialvax)

```
County Demographic Data Editing
```{r, results = "hide"}

countydem3 <- countydem2 %>%
  mutate(total_in_use_housing = total_housing_units - vacant_housing) %>%
  mutate(renter_housing_prop = (renters/(total_in_use_housing))) %>%
  mutate(owner_housing_prop = (owners/(total_in_use_housing))) %>%
  mutate(renter_owner_ratio = renters/owners)
```
\newpage

#Data Dictionary

```{r, results = "hide"}

str(countydem3)
str(covidvax5)

```

#County Demographics Data Set

Name: county_name
Data type: character
Description: A list of all 58 counties in California.

Name:renters
Data type: integer
Description: Number of renters in each county in California.

Name:owners
Data type: integer
Description: Number of owners in each county in California.

Name:vacant_housing
Data type: integer
Description: Number of empty homes in each county in California.

Name:total_housing_units
Data type: integer
Description: Number of total homes in each county in California.

Name:total_in_use_housing
Data type: integer
Description: Number of total homes currently in use in each county in 
California.

Name:renter_housing_prop
Data type: number
Description: Proportion of renters living in occupied homes in 
each county in California.

Name:owner_housing_prop
Data type: number
Description: Proportion of home owners living in occupied homes in 
each county in California.

Name:renter_owner_ratio
Data type: number
Description: Renter to Home owner ratio for 
each county in California.

\newpage

#Covid Vaccination Rate Data Set

Name:county
Data type: character
Description: A list containing the 58 counties in California.

Name:date
Data type: character
Description:The dates when the data was compiled, ranging from January 5, 2021 to
September 14, 2021.

Name:month
Data type: character
Description: The month in which the testing took place. Created to more easily
combine data from specific dates into monthly variables for each county 
in the future.

Name:county_total_pop_over12
Data type: number
Description: Total population over the age of 12 years old in 
each county in California for a specific month.

Name:county_fully_vaccinated
Data type: number
Description: Total number of people fully vaccinated over the age of 12 in 
each county in California for a specific month.

Name:county_fully_vaccinated
Data type: number
Description: Total number of people fully vaccinated over the age of 12 in 
each county in California for a specific month.

Name:county_partially_vaccinated
Data type: number
Description: Total number of people partially vaccinated over the age of 12 in 
each county in California for a specific month.

Name:county_percent_fullvax
Data type: number
Description: Percentage of people fully vaccinated over the age of 12 over the
total population over 12 years old in each county in California for a specific
month.

Name:county_percent_partialvax
Data type: number
Description: Percentage of people partially vaccinated over the age of 12 over the
total population over 12 years old in each county in California for a specific
month.

\newpage

Descriptive Statistics Tables
```{r}
descstatsdem <- countydem3 %>%
  summarise(total_in_use_housing_range = summary(total_in_use_housing),
            renter_housing_ratio_range = summary(renter_housing_prop),
            owner_housing_ratio_range = summary(owner_housing_prop),
            renter_owner_ratio_range = summary(renter_owner_ratio)) 
rownames(descstatsdem) <- c("Min", "1st Quartile", "Median", "Mean", "3rd Quartile", "Max")
```

```{r}
descstatsdem2 <- countydem3 %>%
  summarise(total_renters = sum(renters),
            total_owners = sum(owners))
rownames(descstatsdem2) <- c("Total Count")

```


```{r}
library(kableExtra)
countydem_kable <- kable(descstatsdem,longtable=T,booktabs=T, 
                         col.names = c("Total in Use Housing", "Renter/Housing Ratio", 
                                       "Owner/Housing Ratio", "Renter/Owner Ratio"), 
                         caption = "Summary of CA Housing Across All Counties in 2021")
countydem_kable
countydem_kable2 <-kable(descstatsdem2,longtable=T,booktabs=T, 
                         col.names = c("Renters", "Owners"), 
                         caption = "Total Number of Renters and Owners in CA in 2021")
countydem_kable2
#does not make sense to visualize covidvax dataset in a kable, will update visualization 
#with graph using ggplot in milestone 4
```
\newpage

# Milestone 4

```{r} 
#Creating a combined data set.

countydem3$county_name <- toupper(countydem3$county_name)


combined <- right_join(countydem3, covidvax5, by=c("county_name" = "county"))
```

```{r}
#Table 

combined <- combined %>% filter(county_name == "SAN FRANCISCO" | 
                                county_name == "LOS ANGELES" | 
                                county_name =="MONTEREY" |
                                county_name == "SANTA BARBARA" |
                                county_name == "YOLO" |
                                county_name == "SIERRA" | 
                                county_name == "NEVADA" | 
                                county_name =="EL DORADO" |
                                county_name == "AMADOR" |
                                county_name == "CALAVERAS") %>%
  select(county_name, month, renter_owner_ratio, county_percent_fullvax)

library(tidyr)

combined1 <- combined %>% pivot_wider(names_from = month, 
                                     values_from = county_percent_fullvax) 

combined1 <- combined1 %>% arrange(desc(renter_owner_ratio))

combined2 <- combined1 %>% mutate_if(is.numeric, round, digits = 3)

combined_kable <- kable(combined2,longtable=T,booktabs=T,
                         col.names = c("County", "Renter/Owner Ratio", 
                                       "Jan", "Feb", "March", "April", 
                                       "May", "June", 
                                       "July", "Aug", "Sep"), 
                         caption = "COVID-19 Vaccination Rate by Renter/Owner 
                       Ratio in Top 5 and Bottom 5 Counties in CA in 2021") %>%
  kable_styling(combined_kable, 
                bootstrap_options = c("striped", "hover"), full_width = F) %>%
  row_spec(1:5, bold = TRUE, color = "#CB3155") %>%
  row_spec(6:10, bold = TRUE, color = "#28A69D") %>%
  footnote(general = "Data in pink corresponds with the Top 5 Renter/Owner
           Ratios and their counties while the data in blue is for the Bottom 5. ")
  
combined_kable
```

This table generated via the kable package shows the vaccination rate at the end
of each month for the time period provided in the dataset
for the top 5 and bottom 5 renter-to-owner ratios and their counties. 
The table is in descending order of the ratios and as a visual guide, we have 
color coded specific rows to help split the top 5 from the bottom 5 counties and
all of their corresponding data. 


```{r}

#Using our "combined" data set, we identify the top 5 and bottom 5 renter to 
#owner ratios and their counties using a slice of the original dataset.
#Then we chose 0.75 as a somewhat "arbitrary" ratio to split the two separate 
#groups based off of a simple visual analysis.

countydem4 <- countydem3 %>% arrange(desc(renter_owner_ratio))

countydem4subset <- countydem4 %>% slice(1:5, 54:58)

countydem4subset$county_name <- factor(countydem4subset$county_name, 
                                       levels = countydem4subset$county_name[order(10:1)])
                       
countydem4subset <- countydem4subset %>% 
  mutate(status = 
         case_when(
                   renter_owner_ratio >= 0.75 ~ "Top 5",
                   renter_owner_ratio < 0.75 ~ "Bottom 5"
                                          ))
countydem4subset$status <- factor(countydem4subset$status, 
                                  levels = c("Top 5", "Bottom 5"))
```




\newpage

```{r}
library(ggplot2)

#Graph 1 creation

countydem4subset %>% ggplot(aes(y= county_name, x=renter_owner_ratio)) + 
  geom_point(aes(color=status)) + 
  labs(x = "Renter to Owner Ratio", 
       title= "Renter-to-Owner Ratios in CA: Top 5 and Bottom 5 Counties ",
       y ="California County", fill ="Rank", 
       color="County Status") + 
  scale_x_continuous(breaks=seq(0,2,0.10)) +
  theme_linedraw()

```
This graph provides a greater visual aid for the disparity of owning a home
and the amount of renters for the top 5 and bottom 5 counties. 
As we would expect, we see a very drastic divide between Sierra county and 
Yolo county and that San Francisco is an outlier in this graph, with an 
extremely high amount of people renting housing compared
to who actually owns homes.

```{r}
#Graph 2 creation
covidvax6top5 <- covidvax5 %>% filter(county == "SAN FRANCISCO" | 
                                      county == "LOS ANGELES" | 
                                      county =="MONTEREY" |
                                      county == "SANTA BARBARA" |
                                      county == "YOLO")

covidvax6top5 <- covidvax6top5 %>% select(-c(county_total_pop_over12, 
                                             county_fully_vaccinated, 
                                             county_partially_vaccinated))

covidvax6bottom5 <- covidvax5 %>% filter(county == "SIERRA" | 
                                         county == "NEVADA" | 
                                         county =="EL DORADO" |
                                         county == "AMADOR" |
                                         county == "CALAVERAS")

covidvax6bottom5 <- covidvax6bottom5 %>% select(-c(county_total_pop_over12, 
                                                   county_fully_vaccinated, 
                                             county_partially_vaccinated))
```


```{r}

covidvax6top5$month <- factor(covidvax6top5$month, 
                                  levels = c("January", "February", "March",
                                             "April", "May", "June", "July",
                                             "August", "September", "October",
                                             "November", "December"))

covidvax6top5$county <- factor(covidvax6top5$county, 
                                  levels = c("SAN FRANCISCO", "LOS ANGELES",
                                             "MONTEREY", "SANTA BARBARA",
                                             "YOLO"))

ggplot(data = covidvax6top5, aes(x=month, y=county_percent_fullvax, 
                                 color = county, group = county)) +
  geom_point() +
  geom_line() +
  labs(x="Month", y="Percentage of Fully Vaccinated Individuals", 
       title="COVID-19 Vaccination Rates: Top 5 Counties by Renter/Owner Ratio",
       color="County") +
  theme_bw()
```
This graph shows the vaccination rates for the top 5 renter-to-owner
ratio counties in California over January to September. 
San Francisco consistently had the highest percentage of fully vaccinated
individuals starting in March and reached an astonishing 80% coverage of its 
"over 12" population. Other counties had a very similar pattern of coverage
during this period.



```{r}
covidvax6bottom5$month <- factor(covidvax6bottom5$month, 
                                  levels = c("January", "February", "March",
                                             "April", "May", "June", "July",
                                             "August", "September", "October",
                                             "November", "December"))

covidvax6bottom5$county <- factor(covidvax6bottom5$county, 
                                  levels = c("SIERRA", "NEVADA",
                                             "EL DORADO", "AMADOR",
                                             "CALAVERAS"))

ggplot(data = covidvax6bottom5, aes(x=month, y=county_percent_fullvax, 
                                    color = county, group = county)) +
  geom_point() +
  geom_line() +
  labs(x="Month", y="Percentage of Fully Vaccinated Individuals", 
       title="COVID-19 Vaccination Rate: Bottom 5 Counties by Renter/Owner Ratio",
       color="County") +
  theme_bw()
```
This graph shows the vaccination rates for the bottom 5 renter-to-owner
ratio counties over January to September. The Sierra line looks very strange,
as we lose coverage. This could potentially be due to low population density and
movement out of the county perhaps. We can only speculate with the given data.

\newpage
