---
title: "COVID-19 Vaccination Project"
author: "Adrian Cornejo and Victor Guillera"
date: "11/30/2021"
output: pdf_document
---
## Problem Statement 

The goal of our study is to observe the impact between renter versus homeowner status (housing tenure) on COVID-19 vaccination rates per CA county as an indicator of SES. There are multiple factors behind housing tenure that may affect the overall status of SES and subsequent vaccination rates such as the number of individuals occupying a residency, mortgage status, household income, or generational wealth, and other demographic information. While renter and homeowner rates are not a perfect representation of SES, this comparison will help us elucidate if the financial status is a potential barrier to vaccination and an adequate indicator of SES per our given datasets. 

We calculated the renter-to-homeowner ratio of every county in CA and compared the top 5 ratios to the bottom 5 ratios, tracking vaccine progress by percentage of fully vaccinated individuals over the age of 12 from January 2021 through September 2021. Two different datasets were used, one is of COVID-19 vaccination progress throughout CA provided by the CDPH following vaccine dosage by zip code from 1/5/21 to 9/14/21, and the other is demographic information in CA across all counties from the US census data for 2014 to 2019. 




## Methods 


### Data Sources - Years/Dates of Data:

COVID-19 Vaccination Status Dataset: This dataset supplied by the CDPH tracked the total number of persons partially and fully vaccinated with a COVID-19 vaccine over the age of 12 by CA county and zip code. The dataset contained weekly totals starting on January 5th, 2021 and ended on September 14th, 2021. 

CA County Demographics Dataset: This dataset contained demographic information of every county in CA from 2021 census data. Variables included: population, population density, race/ethnicity, gender, median age, households, families, housing units, average family size, vacant homes, and renter and owner occupancies. 

### Data Cleaning/Methods Used:

\newpage

```{r, echo=F, message=F}
library(readr)
library(knitr)
library(tibble)
library(dplyr)
library(stringr)
library(ggplot2)
library(kableExtra)

countydem <- read.csv("ca_county_demographics.csv")
covidvax <- read.csv("cov_vax_admin.csv")

#To import the csv files, we used read_csv from base R and will 
#inspect our imported files first before we clean our data set.
#Libraries we loaded in were readr, dplyr, and tibble. We are going to pull 
#variables "persons_fully_vaccinated", "persons_partially_vaccinated", 
#"zip_code_tabulation_area", "age12_plus_population", and "county" from the
#covidvax csv file. We will pull "name", "renter_occ","owner_occ", "hse_units" 
#and "vacant" variables.
```

```{r, echo=F, eval=F}
str(covidvax)
head(covidvax)
nrow(covidvax)
ncol(covidvax)
typeof(covidvax)
class(covidvax)
```

```{r, echo=F, eval=F}
distinct(covidvax, as_of_date, .keep_all = F)

#The structure of the "cov_vax_admin" file is a data frame with 11 columns and 
#65,268 rows. The data frame contains integer, number, and character data types 
#and contains a redacted column that is inaccessible. There are also a total of 
#27 unique dates, each 7 weeks apart (weekly results). 

```

```{r, echo=F, eval=F}
str(countydem)
head(countydem)
nrow(countydem)
ncol(countydem)
typeof(countydem)
class(countydem)

#The structure of "countydem" reveals a data frame containing 
#23 columns and 58 rows with integer, character and number data types.

```


```{r, echo=F} 
#Cleaning the data by removing unnecessary columns, rearranging the data,
#and removing "N/A" data points.

#COVID Vaccination Rates
covidvax2 <- select(covidvax, "county","zip_code_tabulation_area", "as_of_date",
                    "age12_plus_population", "persons_fully_vaccinated",
                    "persons_partially_vaccinated")

names(covidvax2) <- c("county", "zip_code", "date", "total_pop_over_12", 
                      "fully_vaccinated", "partially_vaccinated")


covidvax2[is.na(covidvax2)] <- 0

#Using the function str() we see that the selected varibales in covidvax2 
#are chr, int, chr, num, num 
#We may need to convert the dates to numeric using the as.numeric() function 
#to determine and compare vaccination rate. All other column types should be 
#fine as is. 
#Converting NA cells to 0 we allow ourselves to manipulate and describe entire 
#numeric columns.
```

```{r, echo=F, eval=F}

#Descriptions

distinct(covidvax2, county, .keep_all = F)
distinct(covidvax2, date, .keep_all = F)
distinct(covidvax2, zip_code, .keep_all = F)

covidvax2 %>% group_by(zip_code) %>%
  summarise(fullyvax_range = range(fully_vaccinated), partiallyvax_range = range(partially_vaccinated))

covidvax2 %>% 
  summarise(total_pop_range = range(total_pop_over_12), total_pop_median = median(total_pop_over_12))
```

```{r, echo=F}
#County Data 

#Using the function str() we see that the selected varibales in 
#countydem2 are chr, int, int, int, int. All column types should be fine as is 
#for our study question. Converting NA cells to 0 we allow ourselves to 
#manipulate and describe entire numeric columns. We also changed column names to 
#allow for easier/obvious verbiage.

countydem2 <- select(countydem, "name", "renter_occ","owner_occ", 
                     "vacant", "hse_units")

names(countydem2) <-c("county_name","renters","owners",
                      "vacant_housing", "total_housing_units")

countydem2[is.na(countydem2)] <- 0
```

```{r, echo=F, eval=F}

#Descriptions

distinct(covidvax2, county, .keep_all = F)
distinct(covidvax2, date, .keep_all = F)
distinct(covidvax2, zip_code, .keep_all = F)

```

```{r, echo=F, eval=F}

covidvax2 %>% group_by(zip_code) %>% 
  summarise(fullyvax_range = range(fully_vaccinated), partiallyvax_range = range(partially_vaccinated))

``` 

```{r, echo=F,eval=F}

#Population Range and Median for COVID vaccinations for those over 12 years old.

covidvax2 %>% summarise(total_pop_range = range(total_pop_over_12),
                        total_pop_median = median(total_pop_over_12))

#Using the function str() we see that the selected variables 
#in countydem2 are chr, int, int, int, int 
#All column types should be fine as is for our study question
#Converting NA cells to 0 we allow ourselves to manipulate and describe 
#entire numeric columns.
#Changed column names to easier/obvious verbiage
```


```{r, echo=F, eval=F}

#Descriptions

distinct(countydem2, county_name, .keep_all = F)

countydem2 %>% summarise(renters_mean = mean(renters),
                         owners_mean = mean(owners), 
                         vacant_mean = mean(vacant_housing), total_mean = mean(total_housing_units))

countydem2 %>% summarise(renters_range = range(renters), 
                         owners_range = range(owners), 
                         vacant_range = range(vacant_housing), total_range = range(total_housing_units))
```

```{r, echo=F, eval=F}

#Below, is the minimum, first quartile, median, third quartile, and max values.

fivenum(covidvax2$total_pop_over_12)
fivenum(covidvax2$fully_vaccinated)
fivenum(covidvax2$partially_vaccinated)
```

```{r, echo=F, eval=F}

#Descriptions

#We get the mean & range for all numeric and integer vectors in countydem2 below

distinct(countydem2, county_name, .keep_all = F)

``` 

```{r, echo=F, eval=F}

countydem2 %>%
  summarise(renters_mean = mean(renters), 
            owners_mean = mean(owners), 
            vacant_mean = mean(vacant_housing), 
            total_mean = mean(total_housing_units))

countydem2 %>% 
  summarise(renters_range = range(renters), 
                         owners_range = range(owners), 
                         vacant_range = range(vacant_housing), 
                         total_range = range(total_housing_units))

```


```{r, echo=F, eval=F}

#Below, we get the five number summary for data of interest in countydem2.

fivenum(countydem2$renters)
fivenum(countydem2$owners)
fivenum(countydem2$vacant_housing)
fivenum(countydem2$total_housing_units)

```

```{r, echo=F, eval=F}

#Covid Vaccination by County data manipulation (covidvax files)

unique(covidvax2$county)



unique(covidvax2$date)

sum(is.na(covidvax$county)) 
```

```{r, echo=F}
covidvax2 <- covidvax2 %>% filter(county!="0")
```

```{r, echo=F}

covidvax3 <- covidvax2 %>% 
  mutate(month = case_when(
    str_detect(date, "-01-") == T ~ "January",
    str_detect(date, "-02-") == T ~ "February",
    str_detect(date, "-03-") == T ~ "March",
    str_detect(date, "-04-") == T ~ "April",
    str_detect(date, "-05-") == T ~ "May",
    str_detect(date, "-06-") == T ~ "June",
    str_detect(date, "-07-") == T ~ "July",
    str_detect(date, "-08-") == T ~ "August",
    str_detect(date, "-09-") == T ~ "September"
  ))

```

```{r,eval=F, echo=F}

unique(covidvax3$month)

```
`
```{r, echo=F}

covidvax4 <- covidvax3 %>% 
  group_by(county,month)%>%
  mutate(county_total_pop_over12 = sum(total_pop_over_12)) %>%
  mutate(county_fully_vaccinated = sum(fully_vaccinated)) %>%
  mutate(county_partially_vaccinated = sum(partially_vaccinated))
  

covidvax4 <- covidvax4 %>%
  select(-"zip_code",-"total_pop_over_12",-"fully_vaccinated",
         -"partially_vaccinated")

```

```{r,echo=F, eval=F}
#Ensuring we get the last date for each month's final total vaccinations

unique(covidvax4$date)
```

```{r, echo=F}

covidvax4 <- covidvax4 %>% filter(date == "2021-01-26"| date == "2021-02-23" |
                                    date == "2021-03-30" |
                                    date == "2021-04-27" |
                                    date == "2021-05-25" |
                                    date == "2021-06-29" |
                                    date == "2021-07-27" |
                                    date == "2021-08-31" |
                                    date == "2021-09-14")

```

```{r, echo=F}

#Removing duplicate columns after combining the zip codes for each county
#and making column totals for each . Then we check to ensure we have 58 counties
#for each month.

covidvax5 <- covidvax4 %>% distinct(county,month, .keep_all=T)
```

```{r, eval=F, echo=F}

covidvax5 %>% group_by(month) %>% summarise(n())
```

```{r, echo=F}
covidvax5 <- covidvax5 %>%
mutate(county_percent_fullvax =
(county_fully_vaccinated)/(county_total_pop_over12)) %>%
mutate(county_percent_partialvax =
(county_partially_vaccinated)/(county_total_pop_over12))
```

```{r, eval=F, echo=F}
#Checking that our percentages make sense.
summary(covidvax5$county_percent_fullvax)
summary(covidvax5$county_percent_partialvax)

```

```{r, echo=F}

#County Demographic Data Editing


countydem3 <- countydem2 %>%
  mutate(total_in_use_housing = total_housing_units - vacant_housing) %>%
  mutate(renter_housing_prop = (renters/(total_in_use_housing))) %>%
  mutate(owner_housing_prop = (owners/(total_in_use_housing))) %>%
  mutate(renter_owner_ratio = renters/owners)
```


```{r, eval = F, echo=F}
#Data Dictionary

str(countydem3)
str(covidvax5)

```

```{r,eval=F,echo=F}
##County Demographics Data Set

#Name: county_name
#Data type: character
#Description: A list of all 58 counties in California.

#Name:renters
#Data type: integer
#Description: Number of renters in each county in California.

#Name:owners
#Data type: integer
#Description: Number of owners in each county in California.

#Name:vacant_housing
#Data type: integer
#Description: Number of empty homes in each county in California.

#Name:total_housing_units
#Data type: integer
#Description: Number of total homes in each county in California.

#Name:total_in_use_housing
#Data type: integer
#Description: Number of total homes currently in use in each county in 
#California.

#Name:renter_housing_prop
#Data type: number
#Description: Proportion of renters living in occupied homes in 
#each county in California.

#Name:owner_housing_prop
#Data type: number
#Description: Proportion of home owners living in occupied homes in 
#each county in California.

#Name:renter_owner_ratio
#Data type: number
#Description: Renter to Home owner ratio for 
#each county in California.


##Covid Vaccination Rate Data Set

#Name:county
#Data type: character
#Description: A list containing the 58 counties in California.

#Name:date
#Data type: character
#Description:The dates when the data was compiled, ranging from January 5, 2021 to
#September 14, 2021.

#Name:month
#Data type: character
#Description: The month in which the testing took place. Created to more easily
#combine data from specific dates into monthly variables for each county 
#in the future.

#Name:county_total_pop_over12
#Data type: number
#Description: Total population over the age of 12 years old in 
#each county in California for a specific month.

#Name:county_fully_vaccinated
#Data type: number
#Description: Total number of people fully vaccinated over the age of 12 in 
#each county in California for a specific month.

#Name:county_fully_vaccinated
#Data type: number
#Description: Total number of people fully vaccinated over the age of 12 in 
#each county in California for a specific month.

#Name:county_partially_vaccinated
#Data type: number
#Description: Total number of people partially vaccinated over the age of 12 in 
#each county in California for a specific month.

#Name:county_percent_fullvax
#Data type: number
#Description: Percentage of people fully vaccinated over the age of 12 over the
#total population over 12 years old in each county in California for a specific
#month.

#Name:county_percent_partialvax
#Data type: number
#Description: Percentage of people partially vaccinated over the age of 12 over the
#total population over 12 years old in each county in California for a specific
#month.

```


```{r,echo=F}

#Descriptive Statistics Tables

descstatsdem <- countydem3 %>%
  summarise(total_in_use_housing_range = summary(total_in_use_housing),
            renter_housing_ratio_range = summary(renter_housing_prop),
            owner_housing_ratio_range = summary(owner_housing_prop),
            renter_owner_ratio_range = summary(renter_owner_ratio)) 
rownames(descstatsdem) <- c("Min", "1st Quartile", "Median", "Mean", 
                            "3rd Quartile", "Max")
```

```{r,echo=F}
descstatsdem2 <- countydem3 %>%
  summarise(total_renters = sum(renters),
            total_owners = sum(owners))
rownames(descstatsdem2) <- c("Total Count")

```

## Results

```{r, echo=F}

countydem_kable <- kable(descstatsdem,longtable=T,booktabs=T, 
                         col.names = c("Total in Use Housing", "Renter/Housing Ratio", 
                                       "Owner/Housing Ratio", "Renter/Owner Ratio"), 
                         caption = "Summary of CA Housing Across All Counties in 2021")
countydem_kable

```
Table 1 provides a statistical breakdown of the total in-use housing to
total housing,  renter to in-use housing ratio,  owner to in-use housing ratio,
and the renter to owner ratio for the counties in California in 2021.


```{r, echo=F}
countydem_kable2 <-kable(descstatsdem2,longtable=T,booktabs=T, 
                         col.names = c("Renters", "Owners"), 
                         caption = "Total Number of Renters and Owners in CA in 2021")
countydem_kable2

```
Table 2 provides the total renter and owner households in all of California 
in 2021. There are more owner households than renter households overall.


\newpage



```{r, echo=F} 
#Creating a combined data set.

countydem3$county_name <- toupper(countydem3$county_name)


combined <- right_join(countydem3, covidvax5, by=c("county_name" = "county"))
```

\newpage 
```{r, fig.height=8, echo=F}

countydem4 <- countydem3 %>% arrange(desc(renter_owner_ratio))

countydemcombined <- countydem4

countydem4subset <- countydem4 %>% slice(1:5, 54:58)


countydemcombined$county_name <- reorder(countydemcombined$county_name,
                                         countydemcombined$renter_owner_ratio)


countydemcombined %>% ggplot(aes(y= county_name, x=renter_owner_ratio)) + 
  geom_point() + 
  geom_point(data=countydem4subset, 
             aes(x=renter_owner_ratio,y=county_name), 
             color='red',
             size=3) +
  labs(x = "Renter to Owner Ratio", 
       title= "Renter-to-Owner Ratios in CA: All California Counties ",
       y ="California County", fill ="Rank", 
       color="County Status") + 
  scale_x_continuous(breaks=seq(0,2,0.10)) +
  theme(plot.title = element_text(hjust = 0.5)) 

```
Figure 1: This graph displays the Renter-to-Owner Ratio for every single county
in California. Highlighted in red are the top 5 and bottom 5 ratios. We will use
their corresponding counties' data for further analysis.

\newpage


```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), echo=F}

combined <- combined %>% filter(county_name == "SAN FRANCISCO" | 
                                county_name == "LOS ANGELES" | 
                                county_name =="MONTEREY" |
                                county_name == "SANTA BARBARA" |
                                county_name == "YOLO" |
                                county_name == "SIERRA" | 
                                county_name == "NEVADA" | 
                                county_name =="EL DORADO" |
                                county_name == "AMADOR" |
                                county_name == "CALAVERAS") %>%
  select(county_name, month, renter_owner_ratio, county_percent_fullvax)

library(tidyr)

combined1 <- combined %>% pivot_wider(names_from = month, 
                                     values_from = county_percent_fullvax) 

combined1 <- combined1 %>% arrange(desc(renter_owner_ratio))

combined2 <- combined1 %>% mutate_if(is.numeric, round, digits = 3)

combined_kable <- kable(combined2,longtable=T,booktabs=T,
                         col.names = c("County", "Renter/Owner Ratio", 
                                       "Jan", "Feb", "March", "April", 
                                       "May", "June", 
                                       "July", "Aug", "Sep"), 
                         caption = "COVID-19 Vaccination Rate by Renter/Owner 
                       Ratio in Top 5 and Bottom 5 Counties in CA in 2021") %>%
  kable_styling(full_width = F,
                font_size = 7) %>%
  row_spec(1:5, bold = TRUE, color = "#CB3155") %>%
  row_spec(6:10, bold = TRUE, color = "#28A69D") %>%
  footnote(general_title = "Note.", 
           footnote_as_chunk = TRUE,
           threeparttable = TRUE,
           general  ="Pink =Top 5 counties & Blue = Bottom 5 counties."
           )
```
\newpage
```{r, echo=F}
combined_kable
```

Table 3: This table shows the vaccination rate at the end
of each month for the time period provided in the dataset
for the top 5 and bottom 5 renter-to-owner ratios and their counties.



```{r, echo=F}

#Using our "combined" data set, we identify the top 5 and bottom 5 renter to 
#owner ratios and their counties using a slice of the original dataset.
#Then we chose 0.75 as a somewhat "arbitrary" ratio to split the two separate 
#groups based off of a simple visual analysis.

countydem4subset <- countydem4 %>% slice(1:5, 54:58)

countydem4subset$county_name <- factor(countydem4subset$county_name, 
                                       levels = countydem4subset$county_name[order(10:1)])
                       
countydem4subset <- countydem4subset %>% 
  mutate(status = 
         case_when(
                   renter_owner_ratio >= 0.75 ~ "Top 5",
                   renter_owner_ratio < 0.75 ~ "Bottom 5"
                                          ))
countydem4subset$status <- factor(countydem4subset$status, 
                                  levels = c("Top 5", "Bottom 5"))
```




\newpage

```{r, echo=F}
library(ggplot2)

#Graph 1 creation

countydem4subset %>% ggplot(aes(y= county_name, x=renter_owner_ratio)) + 
  geom_point(aes(color=status)) + 
  labs(x = "Renter to Owner Ratio", 
       title= "Renter-to-Owner Ratios in CA: Top 5 and Bottom 5 Counties ",
       y ="California County", fill ="Rank", 
       color="County Status") + 
  scale_x_continuous(breaks=seq(0,2,0.10)) +
  theme_linedraw()

```
Figure 2: This graph provides a greater visual aid for the disparity of owning
a home and the amount of renters for the top 5 and bottom 5 counties. 



\newpage

```{r, echo=F}
#Graph 2 creation
covidvax6top5 <- covidvax5 %>% filter(county == "SAN FRANCISCO" | 
                                      county == "LOS ANGELES" | 
                                      county =="MONTEREY" |
                                      county == "SANTA BARBARA" |
                                      county == "YOLO")

covidvax6top5 <- covidvax6top5 %>% select(-c(county_total_pop_over12, 
                                             county_fully_vaccinated, 
                                             county_partially_vaccinated))

covidvax6bottom5 <- covidvax5 %>% filter(county == "SIERRA" | 
                                         county == "NEVADA" | 
                                         county =="EL DORADO" |
                                         county == "AMADOR" |
                                         county == "CALAVERAS")

covidvax6bottom5 <- covidvax6bottom5 %>% select(-c(county_total_pop_over12, 
                                                   county_fully_vaccinated, 
                                             county_partially_vaccinated))
```


```{r, echo=F}

covidvax6top5$month <- factor(covidvax6top5$month, 
                                  levels = c("January", "February", "March",
                                             "April", "May", "June", "July",
                                             "August", "September", "October",
                                             "November", "December"))

covidvax6top5$county <- factor(covidvax6top5$county, 
                                  levels = c("SAN FRANCISCO", "LOS ANGELES",
                                             "MONTEREY", "SANTA BARBARA",
                                             "YOLO"))

ggplot(data = covidvax6top5, aes(x=month, y=county_percent_fullvax, 
                                 color = county, group = county)) +
  geom_point() +
  geom_line() +
  labs(x="Month", y="Percentage of Fully Vaccinated Individuals", 
       title="COVID-19 Vaccination Rates: Top 5 Counties by Renter/Owner Ratio",
       color="County") +
 theme(axis.text.x = element_text(angle = 30, hjust =1))
```
Figure 3: This graph shows the vaccination rates for the top 5 renter-to-owner
ratio counties in California over January to September. 
San Francisco consistently had the highest percentage of fully vaccinated
individuals starting in March and reached an astonishing 80% coverage of its 
"over 12" population. Other counties had a very similar pattern of coverage
during this period.

\newpage

```{r, echo=F}
covidvax6bottom5$month <- factor(covidvax6bottom5$month, 
                                  levels = c("January", "February", "March",
                                             "April", "May", "June", "July",
                                             "August", "September", "October",
                                             "November", "December"))

covidvax6bottom5$county <- factor(covidvax6bottom5$county, 
                                  levels = c("SIERRA", "NEVADA",
                                             "EL DORADO", "AMADOR",
                                             "CALAVERAS"))

ggplot(data = covidvax6bottom5, aes(x=month, y=county_percent_fullvax, 
                                    color = county, group = county)) +
  geom_point() +
  geom_line() +
  labs(x="Month", y="Percentage of Fully Vaccinated Individuals", 
       title="COVID-19 Vaccination Rate: Bottom 5 Counties by Renter/Owner Ratio",
       color="County") +
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```
Figure 4: This graph shows the vaccination rates for the bottom 5 renter-to-owner
ratio counties over January to September. The Sierra line looks very strange,
as we lose coverage. This could potentially be due to low population density and
movement out of the county perhaps. We can only speculate with the given data.

\newpage

```{r, echo=F}

#Combined Graph Creation Comparing Top and Bottom 5 Counties and Vaccination Rates

covidvaxcombined <- covidvax5 %>% filter(county %in% "SAN FRANCISCO" | 
                                      county == "LOS ANGELES" | 
                                      county =="MONTEREY" |
                                      county == "SANTA BARBARA" |
                                      county == "YOLO" |
                                        county == "SIERRA" | 
                                         county == "NEVADA" | 
                                         county =="EL DORADO" |
                                         county == "AMADOR" |
                                         county == "CALAVERAS")

covidvaxcombined$county <- factor(covidvaxcombined$county, 
                                  levels = c("SAN FRANCISCO", "LOS ANGELES",
                                             "MONTEREY", "SANTA BARBARA",
                                             "YOLO", "SIERRA", "NEVADA",
                                             "EL DORADO", "AMADOR",
                                             "CALAVERAS"))


covidvaxcombined <- covidvaxcombined %>% select(-c(county_total_pop_over12, 
                                             county_fully_vaccinated, 
                                             county_partially_vaccinated))


covidvaxcombined$month <- factor(covidvaxcombined$month, 
                                  levels = c("January", "February", "March",
                                             "April", "May", "June", "July",
                                             "August", "September", "October",
                                             "November", "December"))
```


```{r, echo=F}
covidvaxcombined <- covidvaxcombined %>% 
  mutate(status = 
           case_when(
             county %in% c("SAN FRANCISCO","LOS ANGELES","MONTEREY", "SANTA BARBARA",
                           "YOLO") ~ "Top 5",
             TRUE ~ "Bottom 5"
           ))

covidvaxcombined$status <- factor(covidvaxcombined$status, 
                                  levels = c("Top 5", "Bottom 5"))

```

```{r, fig.height=7, echo=F}

ggplot(data = covidvaxcombined, aes(x=month, y=county_percent_fullvax, 
                                    color = county, group = county)) +
  geom_point(aes(shape = status)) +
  geom_line() +
  labs(x="Month", y="Percentage of Fully Vaccinated Individuals", 
       title="COVID-19 Vaccination Rate: All Counties by Renter/Owner Ratio",
       color="County", shape = " Ratio Rank") +
  scale_y_continuous(breaks=seq(0,1,0.05)) +
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```
Figure 5: This graph combines the data from Figure 3 and Figure 4 for an easier
comparison of vaccination rates over time for each of the top and bottom 5 
counties. 
\newpage

## Discussion

We see in the figures above, that overall, those counties with the highest renter-to-
owner ratios also had the highest vaccination rates for the January-September 
time period compared to the bottom five renter-to-owner ratio counties. We start to
see a divide between the groups after April during the initial vaccine rollout. 
As the data shows, it appears as if being in a county with more home owners is 
more associated with being unvaccinated. 

This goes against our hypothesis that homeowner/renter rates would be a 
good predictor for SES and therefore higher vaccination rates. Owning versus
renting a home involves many SES factors in and of itself, so we can see how
confounding can easily skew our results in this case. Renting in big cities like
Los Angeles and San Francisco can be very expensive, and it can be interpreted 
as showing high SES in some cases for certain counties. It is also important to note the 
political divide in the country and its impact on vaccinations. Bigger cities
like Los Angeles and San Francisco tend to be liberal and more pro vaccine
compared to more conservative rural counties (like the bottom 5, which are all in the Sierra 
Nevada region).


```{r, echo=F} 
#Figuring out what happened to Sierra

covidvaxsierra <- covidvax2 %>% filter(county=="SIERRA")

distinct(covidvaxsierra, zip_code)

#Data for zip code 96118 missing for 1/19/2021, but at this point every county was at zero so i doubt it made a difference. No other data was missing in terms of zip code representation for each period.

covidvaxsierra2 <- covidvaxsierra %>% 
  mutate(month = case_when(
    str_detect(date, "-01-") == T ~ "January",
    str_detect(date, "-02-") == T ~ "February",
    str_detect(date, "-03-") == T ~ "March",
    str_detect(date, "-04-") == T ~ "April",
    str_detect(date, "-05-") == T ~ "May",
    str_detect(date, "-06-") == T ~ "June",
    str_detect(date, "-07-") == T ~ "July",
    str_detect(date, "-08-") == T ~ "August",
    str_detect(date, "-09-") == T ~ "September"
  ))

covidvaxsierra2 <- covidvaxsierra2 %>% 
  group_by(zip_code,month)%>%
  mutate(county_total_pop_over12 = sum(total_pop_over_12)) %>%
  mutate(county_fully_vaccinated = sum(fully_vaccinated)) %>%
  mutate(county_partially_vaccinated = sum(partially_vaccinated))

covidvaxsierra2 <- covidvaxsierra2 %>%
  select(-"total_pop_over_12",-"fully_vaccinated",
         -"partially_vaccinated")

covidvaxsierra2 <- covidvaxsierra2 %>% mutate(percentage = county_fully_vaccinated/county_total_pop_over12)


  
sierrazip96118 <- covidvaxsierra2 %>% filter(zip_code=="96118")

sierrazip95910 <- covidvaxsierra2 %>% filter(zip_code=="95910")

sierrazip96126 <- covidvaxsierra2 %>% filter(zip_code=="96126")

sierrazip95936 <- covidvaxsierra2 %>% filter(zip_code=="95936")

sierrazip96125 <- covidvaxsierra2 %>% filter(zip_code=="96125")

sierrazip96124 <- covidvaxsierra2 %>% filter(zip_code=="96124")

sierrazip95844 <- covidvaxsierra2 %>% filter(zip_code=="95944")

```

```{r, echo=F}


ggplot(sierrazip96125, aes(x=date, y=county_fully_vaccinated)) + geom_point()


#in zip code 95844, 27 people appear fully vaccinated in March and then they disappear.

#no issues for 95910. Nobody ever got vaccinated there.

#in zip code 95936, we peak in March with 613, go down to 518 in April, abruptly sink to zero in May, go back up to 217 in August, then go back to zero.

#in zip code 96118, we see 2769 in June, then a drop to 2291 in July, then an increase to 2953 in August, then a final drop in September at 1234.

# in zip 96124, we peak at 350 in April and drop immediately to zero in May. We then see it go to 171 in august and peak again at 342 in September.

# in zip 96125, we see a peak of 463 in March, a dip to 402 in April, and drop to zero for rest of the year.

#in zip 96126, we see a peak of 791 in June, a dip to 332 in July, goes back to 682 in August, and drops to 173 in September. 
 
```


```{r,echo=F}
rm(covidvaxamador)
#For comparison of data, lets check another small county in the Sierras.
#Lets try Nevada.


covidvaxamador <- covidvax2 %>% filter(county=="AMADOR")

distinct(covidvaxamador, zip_code)

```

```{r,echo=F}

covidvaxamador2 <- covidvaxamador %>% 
  mutate(month = case_when(
    str_detect(date, "-01-") == T ~ "January",
    str_detect(date, "-02-") == T ~ "February",
    str_detect(date, "-03-") == T ~ "March",
    str_detect(date, "-04-") == T ~ "April",
    str_detect(date, "-05-") == T ~ "May",
    str_detect(date, "-06-") == T ~ "June",
    str_detect(date, "-07-") == T ~ "July",
    str_detect(date, "-08-") == T ~ "August",
    str_detect(date, "-09-") == T ~ "September"
  ))

covidvaxamador2 <- covidvaxamador2 %>% 
  group_by(zip_code,month)%>%
  mutate(county_total_pop_over12 = sum(total_pop_over_12)) %>%
  mutate(county_fully_vaccinated = sum(fully_vaccinated)) %>%
  mutate(county_partially_vaccinated = sum(partially_vaccinated))

covidvaxamador2 <- covidvaxamador2 %>%
  select(-"total_pop_over_12",-"fully_vaccinated",
         -"partially_vaccinated")

covidvaxamador2 <- covidvaxamador2 %>% mutate(percentage = county_fully_vaccinated/county_total_pop_over12)

  
amadorzip95669 <- covidvaxamador2 %>% filter(zip_code=="95669")

amadorzip95699 <- covidvaxamador2 %>% filter(zip_code=="95699")

amadorzip95665 <- covidvaxamador2 %>% filter(zip_code=="95665")

amadorzip95640 <- covidvaxamador2 %>% filter(zip_code=="95640")

amadorzip95666 <- covidvaxamador2 %>% filter(zip_code=="95666")

amadorzip95601 <- covidvaxamador2 %>% filter(zip_code=="95601")

amadorzip95646 <- covidvaxamador2 %>% filter(zip_code=="95646")

amadorzip95689 <- covidvaxamador2 %>% filter(zip_code=="95689")

amadorzip95675 <- covidvaxamador2 %>% filter(zip_code=="95675")

amadorzip95629 <- covidvaxamador2 %>% filter(zip_code=="95629")

amadorzip95642 <- covidvaxamador2 %>% filter(zip_code=="95642")

amadorzip95685 <- covidvaxamador2 %>% filter(zip_code=="95685")


#Overall, the rates for Amador's each zip codes varies a bit in terms of total counts and saw #dips, but their percentages always increased and were cumulative. The percentages for Sierra #still saw major disruptions in some zip codes. Its important to note that Amador did see #percentages over 100%, so the way in which data was collected is a mystery to us and how and #why a county's over 12 population would fluctuate so wildly is unknown to us. This is beyond #our control. So we attempt to aggregate the total zip codes by the end of each month to get a #cumulative result for each county to look at the data this way. One possibility is that #counties that were filtered out for having "NA" were mostly Sierra counties or data that had #zeros was intially N/A, so unknown due to lack of reporting so perhaps there were numbers there #but our data processing method of changing all N/A to zero eliminated that. Sierra is very #remote and the smallest county in California, so this is very possible as well.

sum(is.na(covidvax$county)) 

covidvaxlostdata <- covidvax %>% filter(zip_code_tabulation_area == c("96118", "95910",
                                                 "96126","95936", "96125",
                                                 "96124","95944"))




```



