---
title: "COVID Vaccination Project"
author: "Adrian Cornejo and Victor Guillera"
date: "11/4/2021"
output: pdf_document
---

# Milestone 2

Section 1 

This data set will relate to our question if being a homeowner status and 
available housing impacts vaccination rates per CA county. While not a perfect 
representation of SES of cases, this data set will help us elucidate if there is
a trend via linear regression between the homeowner-related to renter rates and
the vaccination rate per county.

We have two data sets, one is of COVID-19 vaccination progress throughout 
CA provided by the CDPH following vaccine dosage by zip code from 1/5/21 
to about present 9/14/21, and the other is demographic info in CA across 
all counties from the US census data for 2014 to 2019.
Below, we read in our libraries, data sets, explore our data sets, 
and then proceed to clean the data sets.

```{r}
library(readr)
library(knitr)
library(tibble)
library(dplyr)
library(stringr)
countydem <- read.csv("ca_county_demographics.csv")
covidvax <- read.csv("cov_vax_admin.csv")

```

\newpage

To import the csv files, we used read_csv from base R and will 
inspect our imported files first before we clean our data set.
Libraries we loaded in were readr, dplyr, and tibble. We are going to pull 
variables "persons_fully_vaccinated", "persons_partially_vaccinated", 
"zip_code_tabulation_area", "age12_plus_population", and "county" from the
covidvax csv file. We will pull "name", "renter_occ","owner_occ", "hse_units" 
and "vacant" variables.
```{r, eval=F}
str(covidvax)
head(covidvax)
nrow(covidvax)
ncol(covidvax)
typeof(covidvax)
class(covidvax)
```

```{r, eval=F}
distinct(covidvax, as_of_date, .keep_all = F)
```
The structure of the "cov_vax_admin" file is a data frame with 11 columns and 
65,268 rows. The data frame contains integer, number, and character data types 
and contains a redacted column that is inaccessible. There are also a total of 
27 unique dates, each 7 weeks apart (weekly results). 

```{r, eval=F}
str(countydem)
head(countydem)
nrow(countydem)
ncol(countydem)
typeof(countydem)
class(countydem)
```
The structure of "countydem" reveals a data frame containing 
23 columns and 58 rows with integer, character and number data types.


\newpage

Cleaning the data by removing unnecessary columns, rearranging the data,
and removing "N/A" data points.
```{r} 

#COVID Vaccination Rates Dataset
covidvax2 <- select(covidvax, "county","zip_code_tabulation_area", "as_of_date",
                    "age12_plus_population", "persons_fully_vaccinated",
                    "persons_partially_vaccinated")

names(covidvax2) <- c("county", "zip_code", "date", "total_pop_over_12", 
                      "fully_vaccinated", "partially_vaccinated")


covidvax2[is.na(covidvax2)] <- 0
```
Using the function str() we see that the selected varibales in covidvax2 
are chr, int, chr, num, num 
We may need to convert the dates to numeric using the as.numeric() function 
to determine and compare vaccination rate. All other column types should be 
fine as is. 
Converting NA cells to 0 we allow ourselves to manipulate and describe entire 
numeric columns.

```{r}
#Descriptions
#distinct(covidvax2, county, .keep_all = F)
#distinct(covidvax2, date, .keep_all = F)
#distinct(covidvax2, zip_code, .keep_all = F)

covidvax2 %>% group_by(zip_code) %>%
  summarise(fullyvax_range = range(fully_vaccinated), 
            partiallyvax_range = range(partially_vaccinated))

covidvax2 %>% 
  summarise(total_pop_range = range(total_pop_over_12), 
            total_pop_median = median(total_pop_over_12))
```

\newpage
```{r}
#County Demographics Dataset

countydem2 <- select(countydem, "name", "renter_occ","owner_occ", 
                     "vacant", "hse_units")

names(countydem2) <-c("county_name","renters","owners",
                      "vacant_housing", "total_housing_units")

countydem2[is.na(countydem2)] <- 0
``` 
Using the function str() we see that the selected varibales in 
countydem2 are chr, int, int, int, int. All column types should be fine as is 
for our study question. Converting NA cells to 0 we allow ourselves to 
manipulate and describe entire numeric columns. We also changed column names to 
allow for easier/obvious verbiage.

```{r, eval=F}

#Descriptions
#distinct(covidvax2, county, .keep_all = F)
#distinct(covidvax2, date, .keep_all = F)
#distinct(covidvax2, zip_code, .keep_all = F)

```

```{r, eval=F}

covidvax2 %>% group_by(zip_code) %>% 
  summarise(fullyvax_range = range(fully_vaccinated), 
            partiallyvax_range = range(partially_vaccinated))

``` 

\newpage 

Population Range and Median for COVID vaccinations for those over 12 years old.
```{r, eval=F}
covidvax2 %>% summarise(total_pop_range = range(total_pop_over_12),
                        total_pop_median = median(total_pop_over_12))
```
Using the function str() we see that the selected variables 
in countydem2 are chr, int, int, int, int 
All column types should be fine as is for our study question
Converting NA cells to 0 we allow ourselves to manipulate and describe 
entire numeric columns.
Changed column names to easier/obvious verbiage

```{r}
#Descriptions

#distinct(countydem2, county_name, .keep_all = F)

countydem2 %>% summarise(renters_mean = mean(renters),
                         owners_mean = mean(owners), 
                         vacant_mean = mean(vacant_housing), 
                         total_mean = mean(total_housing_units))

countydem2 %>% summarise(renters_range = range(renters), 
                         owners_range = range(owners), 
                         vacant_range = range(vacant_housing), 
                         total_range = range(total_housing_units))
```

\newpage

Below, is the minimum, first quartile, median, third quartile, and max values.
```{r}
fivenum(covidvax2$total_pop_over_12)
fivenum(covidvax2$fully_vaccinated)
fivenum(covidvax2$partially_vaccinated)
```

```{r, eval=F}

#Descriptions

#We get the mean & range for all numeric and integer vectors in countydem2 below

distinct(countydem2, county_name, .keep_all = F)
``` 

```{r}
countydem2 %>%
  summarise(renters_mean = mean(renters), 
            owners_mean = mean(owners), 
            vacant_mean = mean(vacant_housing), 
            total_mean = mean(total_housing_units))

countydem2 %>% 
  summarise(renters_range = range(renters), 
                         owners_range = range(owners), 
                         vacant_range = range(vacant_housing), 
                         total_range = range(total_housing_units))

```

\newpage

Below, we get the five number summary for data of interest in countydem2.
```{r}
fivenum(countydem2$renters)
fivenum(countydem2$owners)
fivenum(countydem2$vacant_housing)
fivenum(countydem2$total_housing_units)

```
\newpage 


## Milestone 3

```{r}
#Covid Vaccination by County, Data Manipulation (covidvax files)

unique(covidvax2$county)

covidvax2 <- covidvax2 %>% filter(county!="0")

unique(covidvax2$date)
```
```{r}

covidvax3 <- covidvax2 %>% 
  mutate(month = case_when(
    str_detect(date, "-01-") == T ~ "January",
    str_detect(date, "-02-") == T ~ "February",
    str_detect(date, "-03-") == T ~ "March",
    str_detect(date, "-04-") == T ~ "April",
    str_detect(date, "-05-") == T ~ "May",
    str_detect(date, "-06-") == T ~ "June",
    str_detect(date, "-07-") == T ~ "July",
    str_detect(date, "-08-") == T ~ "August",
    str_detect(date, "-09-") == T ~ "September"
  ))

unique(covidvax3$month)

```
```{r}

covidvax4 <- covidvax3 %>% 
  group_by(county,month)%>%
  mutate(county_total_pop_over12 = sum(total_pop_over_12)) %>%
  mutate(county_fully_vaccinated = sum(fully_vaccinated)) %>%
  mutate(county_partially_vaccinated = sum(partially_vaccinated))
  

covidvax4 <- covidvax4 %>%
  select(-"zip_code",-"total_pop_over_12",-"fully_vaccinated",
         -"partially_vaccinated")

```

```{r}

#Removing duplicate columns after combining the zip codes for each county
#and making column totals for each . Then we check to ensure we have 58 counties
#for each month.

covidvax5 <- covidvax4 %>% distinct(county,month, .keep_all=T)

covidvax5 %>% group_by(month) %>% summarise(n())
```
```{r}
covidvax5 <- covidvax5 %>%
mutate(county_percent_fullvax =
(county_fully_vaccinated)/(county_total_pop_over12)) %>%
mutate(county_percent_partialvax =
(county_partially_vaccinated)/(county_total_pop_over12))

#Checking that our percentages make sense.
summary(covidvax5$county_percent_fullvax)
summary(covidvax5$county_percent_partialvax)
```
\newpage

```{r}
#County Demographic Data Manipulation (countydem files)

countydem3 <- countydem2 %>%
  mutate(total_in_use_housing = total_housing_units - vacant_housing) %>%
  mutate(renter_housing_prop = (renters/(total_in_use_housing))) %>%
  mutate(owner_housing_prop = (owners/(total_in_use_housing))) %>%
  mutate(renter_owner_ratio = renters/owners)
```

\newpage


Data Dictionary
```{r, eval=F}

str(countydem3)
str(covidvax5)

```

County Demographics Data Set:

Name: county_name
Data type: character
Description: A list of all 58 counties in California.

Name:renters
Data type: integer
Description: Number of renters in each county in California.

Name:owners
Data type: integer
Description: Number of owners in each county in California.

Name:vacant_housing
Data type: integer
Description: Number of empty homes in each county in California.

Name:total_housing_units
Data type: integer
Description: Number of total homes in each county in California.

Name:total_in_use_housing
Data type: integer
Description: Number of total homes currently in use in each county in 
California.

Name:renter_housing_prop
Data type: number
Description: Proportion of renters living in occupied homes in 
each county in California.

Name:owner_housing_prop
Data type: number
Description: Proportion of home owners living in occupied homes in 
each county in California.

Name:renter_owner_ratio
Data type: number
Description: Renter to Home owner ratio for 
each county in California.

\newpage

Covid Vaccination Rate Data Set:

Name:county
Data type: character
Description: A list containing the 58 counties in California.

Name:date
Data type: character
Description:The dates when the data was compiled, ranging from January 5, 2021 to
September 14, 2021.

Name:month
Data type: character
Description: The month in which the testing took place. Created to more easily
combine data from specific dates into monthly variables for each county 
in the future.

Name:county_total_pop_over12
Data type: number
Description: Total population over the age of 12 years old in 
each county in California for a specific month.

Name:county_fully_vaccinated
Data type: number
Description: Total number of people fully vaccinated over the age of 12 in 
each county in California for a specific month.

Name:county_fully_vaccinated
Data type: number
Description: Total number of people fully vaccinated over the age of 12 in 
each county in California for a specific month.

Name:county_partially_vaccinated
Data type: number
Description: Total number of people partially vaccinated over the age of 12 in 
each county in California for a specific month.

Name:county_percent_fullvax
Data type: number
Description: Percentage of people fully vaccinated over the age of 12 over the
total population over 12 years old in each county in California for a specific
month.

Name:county_percent_partialvax
Data type: number
Description: Percentage of people partially vaccinated over the age of 12 over the
total population over 12 years old in each county in California for a specific
month.

\newpage

Descriptive Statistics Tables
```{r}
countydem4 <- countydem3 %>%
  summarise(total_in_use_housing_range = summary(total_in_use_housing),
            renter_housing_ratio_range = summary(renter_housing_prop),
            owner_housing_ratio_range = summary(owner_housing_prop),
            renter_owner_ratio_range = summary(renter_owner_ratio)) 

rownames(countydem4) <- c("Min", "1st Quartile", "Median", "Mean", "3rd Quartile", "Max")
```

```{r}
countydem5 <- countydem3 %>%
  summarise(total_renters = sum(renters),
            total_owners = sum(owners))

rownames(countydem5) <- c("Total Count")
```


```{r}
library(kableExtra)

countydem_kable <- kable(countydem4,longtable=T,booktabs=T, 
                         col.names = c("Total in Use Housing", "Renter/Housing Ratio", 
                                       "Owner/Housing Ratio", "Renter/Owner Ratio"), 
                         caption = "Summary of CA Housing Across All Counties in 2021")
countydem_kable

countydem_kable2 <-kable(countydem5,longtable=T,booktabs=T, 
                         col.names = c("Renters", "Owners"), 
                         caption = "Total Number of Renters and Owners in CA in 2021")
countydem_kable2
```

