---
title: "COVID Vaccination Project"
author: "Adrian Cornejo and Victor Guillera"
date: "9/30/2021"
output: pdf_document
---

Section 1 

This data set will relate to our question if being a homeowner status and 
available housing impacts vaccination rates per CA county. While not a perfect 
representation of SES of cases, this data set will help us elucidate if there is
a trend via linear regression between the homeowner-related to renter rates and
the vaccination rate per county.

We have two data sets, one is of COVID-19 vaccination progress throughout 
CA provided by the CDPH following vaccine dosage by zip code from 1/5/21 
to about present 9/14/21, and the other is demographic info in CA across 
all counties. Below, we read in our libraries, data sets, explore our data sets, 
and then proceed to clean the data sets.



```{r}
library(readr)
library(knitr)
library(tibble)
library(dplyr)
countydem <- read.csv("ca_county_demographics.csv")
covidvax <- read.csv("cov_vax_admin.csv")
```

To import the csv files, we used read_csv from base R and will 
inspect our imported files first before we clean our data set.
Libraries we loaded in were readr, dplyr, and tibble. We are going to pull 
variables "persons_fully_vaccinated", "persons_partially_vaccinated", 
"zip_code_tabulation_area", "age12_plus_population", and "county" from the
covidvax csv file. We will pull "name", "renter_occ","owner_occ", "hse_units" 
and "vacant" variables.
```{r, eval=F}
str(covidvax)
head(covidvax)
nrow(covidvax)
ncol(covidvax)
typeof(covidvax)
class(covidvax)
distinct(covidvax, as_of_date, .keep_all = F)
```

The structure of the "cov_vax_admin" file is a data frame with 11 columns and 65,268 rows. The data frame contains integer, number, and character data types and contains a redacted column that is inaccessible. There are also a total of 27 unique dates, each 7 weeks apart (weekly results). 


```{r, eval=F}
str(countydem)
head(countydem)
nrow(countydem)
ncol(countydem)
typeof(countydem)
class(countydem)
```
The structure of "countydem" reveals a data frame containing 
23 columns and 58 rows with integer, character and number data types.



```{r} 
#Cleaning the data by removing unnecessary columns, rearranging the data,
#and removing "N/A" data points.

#COVID Vaccination Rates
covidvax2 <- select(covidvax, "county","zip_code_tabulation_area", "as_of_date",
                    "age12_plus_population", "persons_fully_vaccinated",
                    "persons_partially_vaccinated")

names(covidvax2) <- c("county", "zip_code", "date", "total_pop_over_12", 
                      "fully_vaccinated", "partially vaccinated")


covidvax2[is.na(covidvax2)] <- 0

#County Data 

countydem2 <- select(countydem, "name", "renter_occ","owner_occ", 
                     "vacant", "hse_units")

names(countydem2) <-c("county_name","renters","owners",
                      "vacant_housing", "total_housing_units")

countydem2[is.na(countydem2)] <- 0

```
