---
title: "COVID Vaccination Project"
author: "Adrian Cornejo and Victor Guillera"
date: "9/30/2021"
output: pdf_document
---

Section 1 

This data set will relate to our question if being a homeowner impacts 
vaccination rates per CA county. While not a perfect representation of 
SES of cases, this data set will help us elucidate if there is a trend 
via linear regression between the homeowner to renter rate as well as the 
vaccination rate per county.

We have two datasets, one is of COVID-19 vaccination progress throughout 
CA provided by the CDPH following vaccine doseage by zip code from 1/5/21 
to about present 9/14/21, and the other is demographic info in CA across 
all counties. The code chunk below provides the range of dates for the vaccine 
progress dataset.

```{r}
distinct(covidvax, as_of_date, .keep_all = F)

```

```{r}
library(readr)
library(tibble)
library(dplyr)
countydem <- read.csv("ca_county_demographics.csv")
covidvax <- read.csv("cov_vax_admin.csv")
```
To import the csv files, we used read.csv from the readr package and will 
inspect our imported files first before we clean our data set.
Libraries we loaded in were readr, dplyr, and tibble. We are going to pull variables "persons_fully_vaccinated", "persons_partially_vaccinated", "age12_plus_population", and "county" from the covidvax csv file. We will pull "name", "renter_occ","owner_occ", and "vacant" variables.
```{r}
str(covidvax)
head(covidvax)
tail(covidvax)
view(covidvax)
```

The structure of the "cov_vax_admin" file is a data frame with 11 columns and 65,268 rows. The data frame contains integer, number, and character data types and contains a redacted column that is inaccessible.


```{r}
str(countydem)
head(countydem)
tail(countydem)
view(countydem)
```
The structure of "countydem" reveals a data frame containing 23 columns and 58 rows with integer, character and number data types.



```{r}
covidvax2 <- select(covidvax, "county","as_of_date", "age12_plus_population", "persons_fully_vaccinated", "persons_partially_vaccinated")


view(covidvax2)

countydem2 <- select(countydem, "name", "renter_occ","owner_occ", "vacant")

view(countydem2)
```